Impact of grids and dynamical cores in CESM2.2 on the surface mass balance of the Greenland Ice Sheet

### 2022-09-06 second_submit/arcticgrids.tex
first commit after review

\begin{keypoints}
\item The CESM2.2 release includes several enhancements to the spectral-element dynamical core, including two Arctic refined mesh configurations.
\item Uniform resolution grids degrade the Greenland Ice Sheet mass balance compared to equivalent resolution latitude-longitude grids.
\item The refined Arctic meshes substantially improve the surface mass balance over all low resolution grids.
\end{keypoints}

@@

Six different configurations, a mixture of grids and atmospheric dynamical cores available in the Community Earth System Model, version 2.2 (CESM2.2), are evaluated for their skill in representing the climate of the Arctic and the surface mass balance of the Greenland Ice Sheet (GrIS). The conventional $1^{\circ}-2^{\circ}$ uniform resolution grids systematically overestimate both accumulation and ablation over the GrIS. Of these conventional grids, the latitude-longitude grids outperform the quasi-uniform unstructured grids because they have more degrees of freedom to represent the GrIS. Two Arctic-refined meshes, with $1/4^{\circ}$ and $1/8^{\circ}$ refinement over Greenland, are documented as newly supported configurations in CESM2.2. The Arctic meshes substantially improve the simulated clouds and precipitation rates in the Arctic. Over Greenland, these meshes skillfully represent accumulation and ablation processes, leading to a more realistic GrIS surface mass balance. As CESM is in the process of transitioning away from conventional latitude-longitude grids, these new Arctic-refined meshes improve the representation of polar processes in CESM by recovering resolution lost in the transition to quasi-uniform grids.

@@

The mass balance of Earth's big ice sheets is crucially important for understanding controls on global sea-level rise. However, the scale of the processes needed to represent ice sheet mass balance is challenging to resolve in conventional Earth System Models. This study evaluates the ability of different grids and atmospheric solvers (i.e, the dynamical core) in CESM2.2 to resolve the surface mass balance (SMB) of the Greenland Ice Sheet. We show that the ongoing transition away from latitude longitude grids, towards quasi-uniform unstructured grids in CESM2.2 leads to a degradation of the simulated SMB. Two variable-resolution grids with enhanced resolution over Greenland are developed and incorporated into the release of CESM2.2, which substantially improves the SMB over the latitude longitude grids.
%[ enter your Plain Language Summary here or delete this section]

@@

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high latitudes, which are among the most sensitive regions on Earth to global and environmental change. GCMs differ vastly in their numerical treatment of polar regions because of the so-called \textit{pole problem} \cite{W2007JMSJ}. The pole problem refers to numerical instability arising from the convergence of meridian lines into polar singularities on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a, hereafter referred to as \textit{lat-lon} grids). Depending on the numerics, methods exist to suppress this instability, and lat-lon grids may be advantageous for polar processes by representing structures with finer resolution than elsewhere in the computational domain. With the recent trend towards globally uniform unstructured grids, any potential benefits of lat-lon grids in polar regions may be lost. In this study, we evaluate a number of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 \cite<CESM2.2;>{CESM2}, including new variable-resolution grids, to understand their impacts on the simulated Arctic climate. We focus specifically on the climate and surface mass balance of the Greenland Ice Sheet.

@@

Dycores built on lat-lon grids have some advantages over unstructured grids. Lat-lon coordinate lines are orthogonal, and aligned with zonally symmetric circulations that characterize many large-scale features of Earth's atmosphere. \citeA{LJTN2010JAMES} has experimented with rotating lat-lon models such that their coordinate lines no longer align with an idealized, zonally balanced circulation. For the finite-volume lat-lon dycore considered in this paper (hereafter \textit{FV}), numerical errors were shown to be largest when the polar singularity is rotated into the baroclinic zone ($45^{\circ}$N latitude), generating spurious wave growth much earlier in the simulation than for other rotation angles. This illustrates the advantages of coordinate surfaces aligned with latitude bands, albeit an extreme example where the polar singularity and the polar filter are also contributing to the spurious wave growth. The unstructured grids all generate spurious baroclinic waves earlier in the simulations than the (unrotated) lat-lon models, although the unstructured model considered in this paper, the spectral-element dycore (hereafter \textit{SE}), holds a balanced zonal flow without spurious wave growth appreciably longer than the rotated FV experiments \cite{LJTN2010JAMES}. And unlike FV, the SE dycore has the same error characteristics regardless of how the grid is rotated.

The polar filter in the FV model impedes efficiency at large processor (CPU) counts because it requires a spectral transform, which have large communication overhead \cite{ST1995GEOS,DetAl2012IJHPCA}. Unstructured grids support quasi-uniform grid spacing globally, and there is no pole problem (e.g., Figure~\ref{fig:uni-grids}c). Conversely, unstructured grids are becoming more common due to their improved performance on massively parallel systems and lack of constraints on grid structure \cite{TTI1997JCP,PL2007JCP,WETAL2013GMD}. This grid flexibility allows for the adoption of variable-resolution grids (e.g., Figure~\ref{fig:vr-grids}; hereafter abbreviated as \textit{VR}), sometimes referred to as regional grid refinement. In principle, grid refinement over polar regions can make up for any loss of resolution in transitioning away from lat-lon grids (e.g., Figure~\ref{fig:vr-grids}).  However, local grid refinement comes at the cost of a smaller CFL-limited time step in the refined region; the CFL-condition --- short for Courant–Friedrichs–Lewy condition --- is a necessary condition for numerical stability when using discrete data in time and space.

We emphasize that the pole problem is a distinctive feature of the dycore in atmospheric models. Polar filters do not directly interfere with the physical parameterizations, nor do they have any bearing on the surface models; e.g., the land model can take full advantage of the greater number of grid cells in polar regions on lat-lon grids. This is particularly relevant for the surface mass balance of the Greenland Ice Sheet (\textit{SMB}; the integrated sum of precipitation and runoff), which relies on hydrological processes represented in the land model.

The SMB of the Greenland Ice Sheet (hereafter \textit{GrIS}) is determined by processes occurring over a range of scales that are difficult to represent in GCMs \cite{P2010CC}. GrIS precipitation is concentrated at the ice-sheet margins, where steep topographic slopes drive orographic precipitation. The truncated topography used by low resolution GCMs enables moisture to penetrate well into the GrIS interior, manifesting as a positive precipitation bias \cite{P2000GPC,VETAL2018TC}. GrIS ablation areas (marginal regions where seasonal melting exceeds the annual mass input from precipitation) are typically less than 100\,km wide and are confined to low-lying areas or regions with low precipitation. These narrow ablation zones are not fully resolved in low-resolution GCMs, and may further degrade the simulated SMB. For example, CESM, version 2.0 (CESM2) underestimates ablation in the northern GrIS, leading to unrealistic ice advance when run with an interactive ice sheet component \cite{LETAL2020JAMES}.

@@

In order to retain the benefits of RCMs in a GCM, \citeA{VETAL2018TC} used the VR capabilities of the SE dycore in CESM, generating a grid where Greenland is represented with $1/4^{\circ}$ resolution, and elsewhere with the more conventional $1^{\circ}$ resolution. The simulated SMB compared favorably to the SMB from RCMs and observations. The VR approach is therefore emerging as a powerful tool for simulating and understanding the GrIS and its response to different forcing scenarios.

@@

\begin{figure}[t]
\begin{center}
%\begin{tabular}{cc}
         \includegraphics[width=100mm]{grid-uniform}
         %\includegraphics[width=50mm]{grid-f19.pdf}&
         %\includegraphics[width=50mm]{grid-f09.pdf}\\
         %\includegraphics[width=50mm]{grid-ne30pg2.pdf}&
         %\includegraphics[width=50mm]{grid-ne30pg3.pdf} \\
%\end{tabular}
\end{center}
\caption{Computational grids for the uniform $1^{\circ}-2^{\circ}$ grids in this study. Grids names after Table~\ref{tbl:table1}, (a) $\tt f19$, (b) $\tt f09$, (c) $\tt ne30pg2$ and (d) $\tt ne30pg3$.}
\label{fig:uni-grids}
\end{figure}

@@

\section{Methods}\label{sec:methods}
\subsection{Dynamical cores}

The atmospheric component of CESM2.2 \cite{CESM2}, the Community Atmosphere Model, version 6.3 \cite<CAM6;>{GETAL2019GRL,CAM63}, supports several different atmospheric dynamical cores. These include dycores on lat-lon grids, such as finite-volume \cite<FV;>{L2004MWR} and Eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, including spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. This study compares the performance of the SE and FV dycores, omitting the EUL and FV3 dycores. CESM2 runs submitted to the Coupled Model Intercomparison Project Phase 6 \cite{CMIP6} used the FV dycore, whereas the SE dycore is often used for global high-resolution simulations \cite<e.g., >{SETAL2014JAMES,BETAL2018CC,IHESP} due to its higher throughput on massively parallel systems \cite{DetAl2012IJHPCA}.

@@

The FV dycore is a hydrostatic model that integrates the equations of motion using a finite-volume discretization on a spherical lat-lon grid \cite{LR1997QJR}. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to an Eulerian reference grid in the vertical \cite{L2004MWR}. Hyperviscous damping is applied to the divergent modes, and is increased in the top few layers (referred to as a \textit{sponge layer}) to prevent undesirable interactions with the model top, such as wave reflection \cite{L2011IJHPC}. A polar filter damps computational instability due to the convergence of meridians, permitting a longer time step. It takes the form of a Fourier filter in the zonal direction, with the damping coefficients increasing monotonically in the meridional direction \cite{ST1995GEOS}. The form of the filter is designed to slow down the propagation of large zonal wave-numbers to satisfy the CFL condition of the shortest resolved wave at some reference latitude.

@@

The SE dycore is a hydrostatic model that integrates the equations of motion using a high-order continuous Galerkin method \cite{TTI1997JCP,TF2010JCP}. The computational domain is a cubed-sphere grid tiled with quadrilateral elements (see Figure~\ref{fig:vr-grids}). Each element contains a fourth-order basis set in each horizontal direction, with the solution defined at the roots of the basis functions, the Gauss-Lobatto-Legendre quadrature points. This results in 16 nodal points per element, with 12 of the points lying on the (shared) element boundary. Communication between elements uses the direct stiffness summation \cite{canuto2007}, which applies a numerical flux to the element boundaries to reconcile overlapping nodal values and produce a continuous global basis set.

@@

In CESM2.2, the SE numerics have been enhanced to mitigate spurious noise over topography. These algorithmic changes are described in \ref{sec:a1}.

The SE dycore supports regional grid refinement via its VR configuration, requiring two enhancements over uniform-resolution setups. First, as the numerical viscosity increases with resolution, explicit hyperviscosity relaxes according to the local element size, reducing in strength  by an order of magnitude per halving of the grid spacing. A tensor-hyperviscosity formulation is used \cite{GetAl2014GMD}, which adjusts the coefficients in two orthogonal directions to more accurately target highly distorted quadrilateral elements. Second, the topography boundary conditions are smoothed in a way that does not excite grid scale modes, and so the NCAR topography software \cite{gmdd-8-4623-2015} has been modified to scale the smoothing radius by the local element size, resulting in rougher topography in the refinement zone.

For SE grids with quasi-uniform grid spacing, the SE tracer transport scheme is replaced with the Conservative Semi-Lagrangian Multi-tracer transport scheme (CSLAM) \cite<>{LTOUNGK2017MWR}. Atmospheric tracers have large, nearly discontinuous horizontal gradients that are difficult to represent with spectral methods, which are prone to oscillatory ``Gibbs-ringing" errors \cite{RW1990QJR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a separate grid from the spectral-element dynamics, dividing each element into $3\times3$ control volumes with quasi-equal area. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the original SE dycore in which the physics are evaluated Gauss-Lobatto-Legendre points \cite{HL2018MWR}.

@@

 \begin{table*}
 \centering
 \scriptsize
 \begin{tabular}{lccccc}
   \hline
   grid name & dycore & $\Delta x_{\mathrm{eq}}$ (km) & $\Delta x_{\mathrm{refine}}$ (km) & $\Delta t_{\mathrm{phys}}$ (s) \\ 
   \hline
   $\tt f19$ & FV & 278 & - &1800 \\
   $\tt f09$ & FV & 139 & - &1800 \\
   $\tt ne30pg2$ & SE-CSLAM & 167 & - & 1800 \\
   $\tt ne30pg3$ & SE-CSLAM & 111 & - & 1800 \\
   $\tt ne30pg3^{*}$ & SE-CSLAM & 111 & - & 450 \\
   $\tt Arctic$ & SE & 111 & 28 & 450 \\
   $\tt Arctic-GrIS$ & SE & 111 & 14 & 225 \\
 \hline
 \end{tabular}
  \caption{Grids and dycores used in this study. $\Delta x_{\mathrm{eq}}$ is the average equatorial grid spacing, $\Delta x_{\mathrm{refine}}$ is the grid spacing in the refined region (if applicable), and $\Delta t_{\mathrm{phys}}$ is the physics time step. FV refers to the finite-volume dycore, SE the spectral-element dycore, and SE-CSLAM the spectral-element dycore with CSLAM tracer advection.  We use the $\tt ne30pg3$ grid for two runs with different values of $\Delta t_{\mathrm{phys}}$.}
 \label{tbl:table1}
 \end{table*}

The accuracy of the simulated surface mass balance is expected to be sensitive to grid resolution. Figure~\ref{fig:grisdx}a shows the average grid spacing over the Greenland Ice Sheet (\textit{GrIS} hereafter) in all six grids in this study. The $\tt ne30pg2$ grid has the coarsest representation with an average grid spacing ($\Delta x$) of $\Delta x=160$ km, and the $\tt Arctic-GrIS$ grid has the highest resolution with an average grid spacing of $\Delta x=14.6$ km, similar to the $11$ km grid spacing of the RACMO2.3 grid. The $\tt ne30pg3$ grid has an average $\Delta x=111.2$ km, substantially coarser than the $\tt f09$ grid, with an average $\Delta x=60$ km. Although $\tt ne30pg3$ and $\tt f09$ have similar average grid spacing over the entire globe, and comparable computational costs, the convergence of meridians on the FV grid enhances the resolution over the GrIS. The $\tt Arctic$ grid has an average grid spacing of $\Delta x=27.8$ km, and is about 10 times more expensive than the $1^{\circ}$ models.  The $\tt Arctic-GrIS$ grid is about twice as expensive as the $\tt Arctic$ grid.

@@

Initial simulations with the $\tt ne30pg3$ SE grid produced weaker shortwave cloud forcing relative to the tuned finite-volume dycore in the standard CESM2 configuration. The SE dycore in CESM2.2 therefore has two CLUBB parameter changes to provide more realistic cloud forcing and top-of-atmosphere radiation balance. We reduced the width of the sub-grid distribution of vertical velocity ($\tt clubb\_gamma=0.308\rightarrow0.270$) and also reduced the strength of the damping for horizontal component of turbulent energy ($\tt clubb\_c14=2.2\rightarrow1.6$) to increase cloudiness. For a description of how CLUBB parameters impact the simulated climate, see \citeA{GETAL2015JAMES}.

>>>>

\begin{keypoints}
\item The CESM2.2 release includes several enhancements to the spectral-element dynamical core, including two Arctic refined mesh configurations.
\item {\color{blue}{Quasi-uniform unstructured}} grids degrade the Greenland Ice Sheet mass balance compared to latitude-longitude grids, {\color{blue}{at the conventional $1^{\circ}$ resolution}}.
\item The refined Arctic meshes substantially improve the surface mass balance over {\color{blue}{conventional grid resolutions}}.
\end{keypoints}

@@

Six different configurations, a mixture of grids and atmospheric dynamical cores available in the Community Earth System Model, version 2.2 (CESM2.2), are evaluated for their skill in representing the climate of the Arctic and the surface mass balance of the Greenland Ice Sheet (GrIS). The conventional $1^{\circ}-2^{\circ}$ resolution grids systematically overestimate both accumulation and ablation over the GrIS. Of these conventional grids, the latitude-longitude grids outperform the quasi-uniform unstructured grids because they have more degrees of freedom to represent the GrIS. Two Arctic-refined meshes, with $1/4^{\circ}$ and $1/8^{\circ}$ refinement over Greenland, are documented as newly supported configurations in CESM2.2. The Arctic meshes substantially improve the simulated clouds and precipitation rates in the Arctic. Over Greenland, these meshes skillfully represent accumulation and ablation processes, leading to a more realistic GrIS surface mass balance. As CESM is in the process of transitioning away from conventional latitude-longitude grids, these new Arctic-refined meshes improve the representation of polar processes in CESM by recovering resolution lost in the transition to quasi-uniform grids, albeit at increased computational cost.

@@

The mass balance of Earth's big ice sheets is crucially important for understanding controls on global sea-level rise. However, the scale of the processes needed to represent ice sheet mass balance is challenging to resolve in conventional Earth System Models. This study evaluates the ability of different grids and atmospheric solvers (i.e, the dynamical core) in CESM2.2 to resolve the surface mass balance (SMB) of the Greenland Ice Sheet. We show that the ongoing transition away from latitude{\color{blue}{-}}longitude grids, towards quasi-uniform unstructured grids in CESM2.2 leads to a degradation of the simulated SMB. Two variable-resolution grids with enhanced resolution over Greenland are developed and incorporated into the release of CESM2.2, which substantially improves the SMB over the latitude longitude grids.
%[ enter your Plain Language Summary here or delete this section]

@@

General Circulation Models (GCMs) are powerful tools for understanding the meteorology and climate of the high latitudes, which are among the most sensitive regions on Earth to global and environmental change. GCMs differ vastly in their numerical treatment of polar regions because of the so-called \textit{pole problem} \cite{W2007JMSJ}. The pole problem refers to numerical instability arising from the convergence of meridian lines into polar singularities on latitude-longitude grids (e.g., Figure~\ref{fig:uni-grids}a, hereafter referred to as \textit{lat-lon} grids). Depending on the numerics, methods exist to suppress this instability, and lat-lon grids may be advantageous for polar processes by representing structures with finer resolution than elsewhere in the computational domain. With the recent trend towards quasi-uniform unstructured grids, any potential benefits of lat-lon grids in polar regions may be lost {\color{blue}{(hereafter, {\textit{quasi-uniform}} refers to approximately isotropic grids, which is not a characteristic of lat-lon grids, which are highly anisotropic)}}. In this study, we evaluate a number of grids and dynamical cores (hereafter referred to as \textit{dycores}) available in the Community Earth System Model, version 2.2 \cite<CESM2.2;>{CESM2}, including new variable-resolution grids, to understand their impacts on the simulated Arctic climate. We focus specifically on the climate and surface mass balance of the Greenland Ice Sheet.

@@

Dycores built on lat-lon grids have some advantages over {\color{blue}{dycores built on}} unstructured grids. Lat-lon coordinate lines are orthogonal, and aligned with zonally symmetric circulations that characterize many large-scale features of Earth's atmosphere. \citeA{LJTN2010JAMES} has experimented with rotating lat-lon models such that their coordinate lines no longer align with an idealized, zonally balanced circulation. For the finite-volume lat-lon dycore considered in this paper (hereafter \textit{FV}), numerical errors were shown to be largest when the polar singularity is rotated into the baroclinic zone ($45^{\circ}$N latitude), generating spurious wave growth much earlier in the simulation than for other rotation angles. This illustrates the advantages of coordinate surfaces aligned with latitude bands, albeit an extreme example where the polar singularity and the polar filter are also contributing to the spurious wave growth. The unstructured grids all generate spurious baroclinic waves earlier in the simulations than the (unrotated) lat-lon models, although the unstructured model considered in this paper, the spectral-element dycore (hereafter \textit{SE}), holds a balanced zonal flow without spurious wave growth appreciably longer than the rotated FV experiments \cite{LJTN2010JAMES}. And unlike FV, the SE dycore has the same error characteristics regardless of how the grid is rotated.

The polar filter in the FV model impedes efficiency at large processor (CPU) counts because it requires a spectral transform, which {\color{blue}{has a}} large communication overhead \cite{ST1995GEOS,DetAl2012IJHPCA}. Unstructured grids support quasi-uniform grid spacing globally, and there is no pole problem (e.g., Figure~\ref{fig:uni-grids}c). {\color{blue}{This is in part why}} unstructured grids are becoming more common; their improved performance on massively parallel systems and lack of constraints on grid structure \cite{TTI1997JCP,PL2007JCP,WETAL2013GMD}. This increased grid flexibility allows for the adoption of variable-resolution grids (e.g., Figure~\ref{fig:vr-grids}; hereafter abbreviated as \textit{VR}), sometimes referred to as regional grid refinement. In principle, grid refinement over polar regions can make up for any loss of resolution in transitioning away from lat-lon grids (e.g., Figure~\ref{fig:vr-grids}).  However, local grid refinement comes at the cost of a smaller CFL-limited time step in the refined region; the CFL-condition --- short for Courant–Friedrichs–Lewy condition --- is a necessary condition for numerical stability when using discrete data in time and space.

We emphasize that the pole problem is a distinctive feature of the dycore in atmospheric models. Polar filters do not directly interfere with the physical parameterizations, nor do they have any bearing on the surface models; e.g., the land model can take full advantage of the greater number of grid cells in polar regions on lat-lon grids. This is particularly relevant for the surface mass balance {\color{blue}{(\textit{SMB}; the integrated sum of precipitation and runoff)}} of the Greenland Ice Sheet, which relies on hydrological processes represented in the land model.

The SMB of the Greenland Ice Sheet (hereafter \textit{GrIS}) is determined by processes occurring over a range of scales that are difficult to represent in GCMs \cite{P2010CC}. GrIS precipitation is concentrated at the ice-sheet margins, where steep topographic slopes drive orographic precipitation. The truncated topography used by low resolution GCMs enables moisture to penetrate well into the GrIS interior, manifesting as a positive precipitation bias \cite{P2000GPC,VETAL2018TC}. GrIS ablation areas {\color{blue}{(marginal regions where the annual SMB is negative)}} are typically less than 100\,km wide and are confined to low-lying areas or regions with low precipitation. These narrow ablation zones are not fully resolved in low-resolution GCMs, and may further degrade the simulated SMB. For example, CESM, version 2.0 (CESM2) underestimates ablation in the northern GrIS, leading to unrealistic ice advance when run with an interactive ice sheet component \cite{LETAL2020JAMES}.

@@

In order to retain the benefits of RCMs in a GCM, \citeA{VETAL2018TC} used the VR capabilities of the SE dycore in CESM, generating {\color{blue}{grids}} where Greenland is represented with {\color{blue}{up to}} $1/4^{\circ}$ resolution, and elsewhere with the more conventional $1^{\circ}$ resolution. The simulated SMB compared favorably to the SMB from RCMs and observations. The VR approach is therefore emerging as a powerful tool for simulating and understanding the GrIS and its response to different forcing scenarios.

@@

\begin{figure}[t]
\begin{center}
%\begin{tabular}{cc}
         \includegraphics[width=100mm]{grid-uniform}
         %\includegraphics[width=50mm]{grid-f19.pdf}&
         %\includegraphics[width=50mm]{grid-f09.pdf}\\
         %\includegraphics[width=50mm]{grid-ne30pg2.pdf}&
         %\includegraphics[width=50mm]{grid-ne30pg3.pdf} \\
%\end{tabular}
\end{center}
\caption{Computational grids for the $1^{\circ}-2^{\circ}$ {\color{blue}{lat-lon and quasi-uniform unstructured}} grids in this study. Grids names after Table~\ref{tbl:table1}, (a) $\tt f19$, (b) $\tt f09$, (c) $\tt ne30pg2$ and (d) $\tt ne30pg3$.}
\label{fig:uni-grids}
\end{figure}

@@

\section{Methods}\label{sec:methods}

{\color{blue}{CESM2.2 is a CMIP6-class \cite<Coupled Model Intercomparison Project Phase 6;>{CMIP6} Earth System Model maintained by the National Center for Atmosphere Research. CESM2.2 contains sub-component models for the atmosphere, land, ocean, sea-ice, and land-ice, and can be configured to run with varying degrees of complexity. All simulations described in this study use an identical transient 1979-1998 Atmospheric Model Inter-comparison Project (AMIP) configuration, with prescribed monthly sea-surface temperature and sea ice following \citeA{CESMSST}. In CESM terminology, AMIP simulations use the $\tt FHIST$ computational set and run out of the box in CESM2.2. The land-ice component is not dynamically active in $\tt FHIST$. However, the surface mass balance is computed by the land model before being passed to the land-ice component; $\tt FHIST$ includes the functionality to simulate the surface mass balance of the Greenland Ice Sheet.}}

\subsection{Dynamical cores}

The atmospheric component of CESM2.2 \cite{CESM2}, the Community Atmosphere Model, version 6.3 \cite<CAM6;>{GETAL2019GRL,CAM63}, supports several different atmospheric dynamical cores. These include dycores on lat-lon grids, such as finite-volume \cite<FV;>{L2004MWR} and Eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, including spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. This study compares the performance of the SE and FV dycores, omitting the EUL and FV3 dycores. CESM2 runs submitted to CMIP6 used the FV dycore, whereas the SE dycore is often used for global high-resolution simulations \cite<e.g., >{SETAL2014JAMES,BETAL2018CC,IHESP} due to its higher throughput on massively parallel systems \cite{DetAl2012IJHPCA}.

@@

The FV dycore integrates the {\color{blue}{hydrostatic}} equations of motion using a finite-volume discretization on a spherical lat-lon grid \cite{LR1997QJR}. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to an Eulerian reference grid in the vertical \cite{L2004MWR}. Hyperviscous damping is applied to the divergent modes, and is increased in the top few layers (referred to as a \textit{sponge layer}) to prevent undesirable interactions with the model top, such as wave reflection \cite{L2011IJHPC}. A polar filter damps computational instability due to the convergence of meridians, permitting a longer time step. It takes the form of a Fourier filter in the zonal direction, with the damping coefficients increasing monotonically in the meridional direction \cite{ST1995GEOS}. The form of the filter is designed to slow down the propagation of large zonal wave-numbers to satisfy the CFL condition of the shortest resolved wave at some reference latitude.

@@

The SE dycore integrates the {\color{blue}{hydrostatic}} equations of motion using a high-order continuous Galerkin method \cite{TTI1997JCP,TF2010JCP}. The computational domain is a cubed-sphere grid tiled with quadrilateral elements (see Figure~\ref{fig:vr-grids}). Each element contains a fourth-order basis set in each horizontal direction, with the solution defined at the roots of the basis functions, the Gauss-Lobatto-Legendre quadrature points. This results in 16 nodal points per element, with 12 of the points lying on the (shared) element boundary. Communication between elements uses the direct stiffness summation \cite{canuto2007}, which applies a numerical flux to the element boundaries to reconcile overlapping nodal values and produce a continuous global basis set.

@@

In CESM2.2, the SE numerics have been enhanced {\color{blue}{relative to the CESM2.0 release}} to mitigate spurious noise over topography. These algorithmic changes are described in \ref{sec:a1}.

The SE dycore supports regional grid refinement via its VR configuration, requiring two enhancements over {\color{blue}{quasi-uniform resolution grids}}. First, as the numerical viscosity increases with resolution, explicit hyperviscosity relaxes according to the local element size, reducing in strength  by an order of magnitude per halving of the grid spacing. A tensor-hyperviscosity formulation is used \cite{GetAl2014GMD}, which adjusts the coefficients in two orthogonal directions to more accurately target highly distorted quadrilateral elements. Second, the topography boundary conditions are smoothed in a way that does not excite grid scale modes, and so the NCAR topography software \cite{gmdd-8-4623-2015} has been modified to scale the smoothing radius by the local element size, resulting in rougher topography in the refinement zone.

When using the SE dycore with quasi-uniform grid spacing, the SE tracer transport scheme is replaced with the Conservative Semi-Lagrangian Multi-tracer transport scheme (CSLAM) \cite<>{LTOUNGK2017MWR}. Atmospheric tracers have large, nearly discontinuous horizontal gradients that are difficult to represent with spectral methods, which are prone to oscillatory ``Gibbs-ringing" errors \cite{RW1990QJR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a separate grid from the spectral-element dynamics, dividing each element into $3\times3$ control volumes with quasi-equal area. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the original SE dycore in which the physics are evaluated at Gauss-Lobatto-Legendre points \cite{HL2018MWR}. {\color{blue}{CSLAM advection is not an available option in the VR configuration, which instead uses the standard SE tracer transport scheme with the physics evaluated at Gauss-Lobatto-Legendre points.}}

Initial simulations with the SE dycore produced weaker shortwave cloud forcing relative to the tuned finite-volume dycore in the standard CESM2 configuration. The SE dycore in CESM2.2 therefore has two CLUBB parameter changes to provide more realistic cloud forcing and top-of-atmosphere radiation balance. We reduced the width of the sub-grid distribution of vertical velocity ($\tt clubb\_gamma=0.308\rightarrow0.270$) and also reduced the strength of the damping for horizontal component of turbulent energy ($\tt clubb\_c14=2.2\rightarrow1.6$) to increase cloudiness. For a description of how CLUBB parameters impact the simulated climate, see \citeA{GETAL2015JAMES}.

@@

 \begin{table*}
 \centering
 \scriptsize
 \begin{tabular}{lcccccc}
   \hline
   grid name & dycore & $\Delta x_{\mathrm{eq}}$ (km) & $\Delta x_{\mathrm{fine}}$ (km) & $\Delta t_{\mathrm{phys}}$ (s) & {\color{blue}{25 nodes, cost}} & {\color{blue}{50 nodes, cost}} \\ 
   \hline
   $\tt f19$ & FV & 278 & - &1800 & {\color{blue}{436.66}} & {\color{blue}{-}} \\
   $\tt f09$ & FV & 139 & - &1800 & {\color{blue}{1534.57}} & {\color{blue}{2024.24}} \\
   $\tt ne30pg2$ & SE-CSLAM & 167 & - & 1800 & {\color{blue}{1497.26}} & {\color{blue}{1683.97}} \\
   $\tt ne30pg3$ & SE-CSLAM & 111 & - & 1800 & {\color{blue}{1890.48}} & {\color{blue}{2090.43}} \\
   $\tt ne30pg3^{*}$ & SE-CSLAM & 111 & - & 450 & {\color{blue}{-}} & {\color{blue}{-}} \\
   $\tt Arctic$ & SE & 111 & 28 & 450 & {\color{blue}{15947.41}} & {\color{blue}{16675.45}} \\
   $\tt Arctic-GrIS$ & SE & 111 & 14 & 225 & {\color{blue}{40305.03}} & {\color{blue}{41036.67}} \\
 \hline
 \end{tabular}
  \caption{Grids and dycores used in this study. $\Delta x_{\mathrm{eq}}$ is the average equatorial grid spacing, $\Delta x_{\mathrm{fine}}$ is the grid spacing in the refined region (if applicable), and $\Delta t_{\mathrm{phys}}$ is the physics time step. FV refers to the finite-volume dycore, SE the spectral-element dycore, and SE-CSLAM the spectral-element dycore with CSLAM tracer advection.  We use the $\tt ne30pg3$ grid for two runs with different values of $\Delta t_{\mathrm{phys}}$. {\color{blue}{The last columns provide the computational costs in core hours per simulated year (CHPSY). The costs are from single month runs using 25 nodes and 50 nodes on the Cheyenne supercomputer \cite{cheyenne}.}}}
 \label{tbl:table1}
 \end{table*}

The accuracy of the simulated surface mass balance is expected to be sensitive to grid resolution. Figure~\ref{fig:grisdx}a shows the average grid spacing over the Greenland Ice Sheet (\textit{GrIS} hereafter) in all six grids, {\color{blue}{as well as two grids pertaining to the Regional Atmospheric Climate Model \cite<RACMO;>{NETAL2018TC,NETAL2019SCIENCE}, which are used for validation purposes in this study (Table~\ref{tbl:table2})}}. The $\tt ne30pg2$ grid has the coarsest representation with an average grid spacing ($\Delta x$) of $\Delta x=160$ km, and the $\tt Arctic-GrIS$ grid has the highest resolution with an average grid spacing of $\Delta x=14.6$ km, similar to the $11$ km grid spacing of the RACMO2.3 grid. The $\tt ne30pg3$ grid has an average $\Delta x=111.2$ km, substantially coarser than the $\tt f09$ grid, with an average $\Delta x=60$ km. Although $\tt ne30pg3$ and $\tt f09$ have similar average grid spacing over the entire globe, and comparable computational costs, the convergence of meridians on the FV grid enhances the resolution over the GrIS. The $\tt Arctic$ grid has an average grid spacing of $\Delta x=27.8$ km.

@@

\subsection{{\color{blue}{Computational costs}}}

{\color{blue}{The last columns of Table~\ref{tbl:table1} provides cost estimates for the different grids and dycores. The costs, expressed as core hours per simulated year (CHPSY), are taken from single month runs of $\tt FHIST$, with no i/o, and using 25 nodes (900 tasks) and 50 nodes (1800 tasks) on the Cheyenne supercomputer \cite{cheyenne}. 25 nodes is on the low side for a typical multi-decadal climate simulation at $1^{\circ}$ resolution, but it's the largest number of tasks that can be supported by the $\tt f19$ grid, and we chose to fix the number of tasks across all grids for the purposes of comparing their costs. We also provide costs using 50 nodes, excluding $\tt f19$, to provide a more practical cost estimate for longer climate integrations. There are probably better approaches for comparing costs across different grids and dycores, e.g., holding fixed the number of grid columns per task, but it is beyond the scope of this study.

%Running with the more common $1^{\circ}$ grid for multi-decadal climate simulations, 25 nodes is on the low side, but it's the largest number of tasks that can be supported by the $\tt f19$ grid, and we chose to fix the number of tasks across all grids for the purposes of comparing their costs.

The cheapest grid is the $\tt f19$ grid at 436.66 CHPSY, as this is the only grid running with $2^{\circ}$ dynamics. The $\tt f09$ grid costs 1534.57 CHPSY using 25 nodes, which is noticeably cheaper than $\tt ne30pg3$ at 1890.48 CHPSY. The $\tt ne30pg2$ grid is 20$\%$ cheaper than the $\tt ne30pg3$ grid, in both the 25 node and 50 node runs, consistent with previous estimates \cite{HR2018JAMES}. The FV model is known to be cheaper than SE at small core counts, whereas SE is more efficient than FV at large core counts due to its improved scalability \cite{DFSSTT2005IJHPC,DetAl2012IJHPCA}. In the more conventional 50 node runs, $\tt f09$ costs are much more similar to $\tt ne30pg3$, due to a 30$\%$ cost increase in $\tt f09$ relative to the 25 node run (Table~\ref{tbl:table1}). The $\tt Arctic$ grid is an order of magnitude more expensive than the lat-lon and quasi-uniform grids, at about 16k CHPSY, whereas the $\tt Arctic-GrIS$ grid is a more than twice that (40k CHPSY). Note that all timing numbers are from runs without threading. The $\tt f09$ grid is the only grid that runs out-of-the-box with threading; holding the number of tasks fixed leads to a $4\%-6.5\%$ reduction in CHPSY compared to runs without threading.}}
